The Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-08 ms.tex
3.1 update

Continuity of the entropy derivative at the stagnation radius where $v = 0$ requires that the temperature at the stagnation radius is given by (eq.~[\ref{eq:first_law}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p \tilde{v}_{w}^{2}}{2} \approx 3.7
\times 10^6\ \, v_{500}^2 \,\,{\rm K},
\label{eq:Tanalytic}
\end{align}
where $v_{500} \equiv \tilde{v}_{w}/(500$ km s$^{-1}$).  Hydrostatic equilibrium furthermore determines the value of the stagnation radius (Appendix \ref{app:rs}, eq.~\ref{eq:rs2main})
\begin{align}
\rs=\frac{G \Mbh}{n \vw^{2}|_{\rs}}\left(\frac{9-\Gamma}{2} \frac{M_{\star}|_{\rs}}{\Mbh} +\frac{7}{2}\right),
\label{eq:stag_analytic}
\end{align}
where $n$ is the power-law slope of the density profile at $r = r_{\rm s}$.

If the SMBH dominates the gravitational potential at the stagnation radius then $v_{w} \gg \sigma$ and hence the stagnation radius resides well inside the sphere of influence, i.e.~$r_{\rm s} < r_{\rm soi}$ (eq.~[\ref{eq:rsoi}]).  In this case equation (\ref{eq:stag_analytic}) simplifies to
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \tilde{v}_{w}^{2}} \approx 6
 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where the numerical estimate assumes $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their eq.~6).  In the opposite limit of weak heating ($v_{w} \ll \sigma$), the stagnation radius moves to large radii $\gg r_{\rm soi}$ or may not exist at all.  This result makes intuitive sense: gas is supplied by stars which are gravitationally bound to the black hole, so outflows occur only if the specific heating rate $\sim \tilde{v}_{w}^{2}$ exceeds the specific gravitational binding energy $\sim \sigma^{2}$.

@@

The accretion rate onto the SMBH is given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have made use of equation (\ref{eq:stag_simple}; assuming that $v_{w} \gg \sigma$) and have divided our results into core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies.  The resulting Eddington ratio $\eddr$ is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$  is the Eddington accretion rate, assuming a radiative efficiency of ten per cent.  Note the sensitive inverse dependence of the accretion rate on the wind heating rate.  
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

The gas density at the stagnation radius $\rho(r_{\rm s})$ is more challenging to estimate accurately.  Using an alternative estimate of $\dot{M}$ as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}|_{r_{\rm s}} = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$ in conjuction with equation (\ref{eq:mdot_analytic}), one finds
\begin{align}
  &\dot{M}\sim\frac{(4 \pi/3) \rs^3 \rho|_{r_{\rm s}}}{\tff|_{r_{\rm s}}},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho|_{r_{\rm s}}\approx
  \begin{cases}
    5 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core},\\
    9 \times 10^{-26}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp},
  \end{cases}
  \label{eq:rhors}
\end{align}
again using equation (\ref{eq:rs_simple}) for the stagnation radius.

@@

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) may be compared to the traditional Bondi rate for accretion onto a point source from an external medium of specified density and temperature (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_b^2 \rho(r_b) v_{\rm ff}(r_b),
\label{eq:bondi}
\end{align}
where $r_b \equiv GM/c_{s}^{2}$ is the Bondi radius, $v_{\rm ff} = (GM_{\bullet}/r)^{1/2}$ and $\lambda$ is a parameter of order unity.  By rewriting equation (\ref{eq:mdot_gas}) as $\dot{M}\simeq 4\pi \rs^2 \rho|_{r_{\rm s}} v_{\rm ff}|_{r_{\rm s}}/3$,
one notices a similarity between equation (\ref{eq:mdot_gas}) and the Bondi formula (eq.~[\ref{eq:bondi}]) if $\rb$ is replaced by $\rs$.  When the stagnation radius resides interior to the sphere of influence we indeed have (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}} = \frac{14}{3}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where the second equality makes use of equation (\ref{eq:Tanalytic}).

>>>>

Continuity of the entropy derivative at the stagnation radius where $v = 0$ requires that the temperature at the stagnation radius is given by (eq.~[\ref{eq:first_law}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p \tilde{v}_{w}^{2}}{2k} \approx 3.7
\times 10^6\ \, v_{500}^2 \,\,{\rm K},
\label{eq:Tanalytic}
\end{align}
where $v_{500} \equiv \tilde{v}_{w}/(500$ km s$^{-1}$).  Hydrostatic equilibrium furthermore determines the value of the stagnation radius (Appendix \ref{app:rs}, eq.~\ref{eq:rs2main})
\begin{align}
\rs=\frac{G \Mbh}{n \vw^{2}|_{\rs}}\left(\frac{9-\Gamma}{2} \frac{M_{\star}|_{\rs}}{\Mbh} +\frac{7}{2}\right),
\label{eq:stag_analytic}
\end{align}
where $n \equiv -d{\rm ln}\rho/d{\rm ln\,r}|_{r_{\rm s}}$ is the density power-law slope at $r = r_{\rm s}$.

If the SMBH dominates the gravitational potential at the stagnation radius, then $v_{w} \gg \sigma$ and hence the stagnation radius resides well inside the sphere of influence ($r_{\rm s} < r_{\rm soi}$; eq.~[\ref{eq:rsoi}]).  In this limit equation (\ref{eq:stag_analytic}) simplifies to
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \tilde{v}_{w}^{2}} \approx 6
 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where for the numerical estimate we adopt $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on more heuristic grounds (their eq.~6).

In the opposite limit of weak heating ($v_{w} \ll \sigma$), the stagnation radius moves to large radii $\gg r_{\rm soi}$ or may not exist at all (formally implying that the entire ISM of the galaxy is accreting).  This result makes intuitive sense: gas is supplied to the nuclear by stars which are gravitationally bound to the black hole, so outflows are possible only if the specific heating rate $\sim \tilde{v}_{w}^{2}$ exceeds the specific gravitational binding energy $\sim \sigma^{2}$.

@@

The accretion rate onto the SMBH is given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have made use of equation (\ref{eq:stag_simple}), i.e. assuming that $v_{w} \gg \sigma$, and we have divided our results into core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies.  The resulting Eddington ratio $\dot{M}/\dot{M}_{\rm edd}$ is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$  is the Eddington accretion rate, assuming a radiative efficiency of ten per cent.  Note the sensitive inverse dependence of the accretion rate on the wind heating rate.  
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

The gas density at the stagnation radius $\rho|_{r_{\rm s}}$ is more challenging to estimate accurately.  Using an alternative estimate of $\dot{M}$ as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}|_{r_{\rm s}} = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$,
\begin{align}
  &\dot{M}\sim\frac{(4 \pi/3) \rs^3 \rho|_{r_{\rm s}}}{\tff|_{r_{\rm s}}},
  \label{eq:mdot_gas}
\end{align}
 in conjuction with equations (\ref{eq:mdot_analytic}) and (\ref{eq:rs_simple}), one derives
\begin{align}
  \rho|_{r_{\rm s}}\approx
  \begin{cases}
    5 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core},\\
    9 \times 10^{-26}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}.
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) can be compared to the traditional Bondi rate for accretion onto a point source from an external medium of specified density and temperature (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_{\rm B}^2 \rho|_{r_{\rm B}}v_{\rm ff}|_{r_{\rm B}},
\label{eq:bondi}
\end{align}
where $r_{\rm B} \equiv GM/c_{\rm s,ad}^{2}$ is the Bondi radius, $c_{\rm s,ad} = \sqrt{\gamma kT/\mu m_p}$ is the adiabatic sound speed, $v_{\rm ff}|_{r_{\rm B}} = r_{\rm B}/t_{\rm ff}|_{r_{\rm B}} = (GM_{\bullet}/r)^{1/2}$ and $\lambda$ is a parameter of order unity.  One observes a similarity between equation (\ref{eq:mdot_gas}) and the Bondi formula (eq.~[\ref{eq:bondi}]) if $R_{\rm B}$ is replaced by $\rs$.  When the stagnation radius resides interior to the sphere of influence we indeed have (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \approx \frac{7}{6n}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where the second equality makes use of equation (\ref{eq:Tanalytic}).  {\bf BDM: please check, as I found something different than what was here before}.
